---
title: "An Introduction to the conclus Package"
output: rmarkdown::html_vignette
fig_width: 8
fit_height: 6
vignette: >
  %\VignetteIndexEntry{An Introduction to the conclus Package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
set.seed(19561013)
library(knitr)
opts_chunk$set(out.extra = 'style="display:block; margin: auto"',
               fig.align = "center", dev = "png")
```

## Introduction ##

_Cluster analysis_, sometimes known as _unsupervised learning_ is a common
statistical task. In short, it involves taking several measurements on each of
several objects, and attempting to find groups (clusters) into which objects
naturally fall. Objects in the same cluster should be more similar to each
other than to those in other clusters. Where it differs from more
common statistical classification tasks, or _supervised learning_, is that
we do not know in advance how many clusters exist (and there might be only 1!).
We therefore use algorithms to guide us.

A popular book-length treatment of the subject is _Finding Groups in Data_
by Kaufman and Rousseeuw (Wiley, 1990, 2005). Unfortunately, that book appears
to be out of print and second hand copies are going at prohibitive prices. For an
alternative introduction, see Chapter 14 of
[Hastie, Tibshirani and Friedman](https://statweb.stanford.edu/~tibs/ElemStatLearn/)
(Springer, 2009).

A difficult in practice is that at some point we need to make a decision about
how many clusters we believe there to be (or at least suggest a small number
of candidates). Tools such as the _gap statistic_ (see the help file for
`clusGap` in the **cluster** package in R) can help with this, but additional
tools are welcome.

A weakness of many clustering algorithms is that they will tend to produce
quite different looking results from similar datasets: they are said to be
_unstable_. Similar issues arise with classification trees, for example.

For unstable algorithms in general, a common approach is to take random samples
from the available data, fit the model to each sample, and then average over
the outputs of the models.
[Breiman named this approach 'Bootstrap aggregating'](https://www.stat.berkeley.edu/~breiman/bagging.pdf)
or _bagging_ for short.

The original bagging idea involved repeatedly taking samples, with replacement,
of the same size as the original data. Various authors have since discovered that
taking random subsamples of 50% (without replacement) leads to similar
results at lower computational expense. [Stability selection](https://arxiv.org/pdf/0809.2932.pdf)
is just one example of this idea in practice.

Given the success of bagging and subsampling in classification, regression and
variable selection, it will be no surprise that similar ideas work in clustering.

## Consensus clustering ##
_Consensus clustering_ was introduced by [Monti et al in 2003](http://www.cs.utexas.edu/users/ml/biodm/papers/MLJ-biodm2.pdf).
The method amounts to the following:

* Take a random subsample (say 50%) of the data;
* Run a clustering algorithm assuming there to be k=2 clusters;
* Save the class memberships of the observations;
* Repeat lots of times;
* Repeat with k=3, 4, ..., K clusters.

The results can then be summarized in order to provide insight on:

* Whether any genuine clusters appear to exist;
* How similar are items are within each cluster;
* What is the best value of k (or a small number of candidate values of k).

Monti et al suggest various summaries and plots to address these questions.

Note that no particular clustering algorithm is specified: the user can use
their own favourite, or try several algorithms to see what different properties
they exhibit on the data to hand.

## The **conclus** package ##
The **conclus** package for [R](https://www.r-project.org/) is an implementation
of consensus clustering that has the following features:

* The user can pass in their own clustering function;
* Various summaries and plots can be easily produced;
* The user interface will seem natural to most R users;
* The code forces the user to think about distance metric (the pain will do you good!);
* Additional information is easily extracted from the returned object.

### Examples ###
Firstly, we'll illustrate the main features of **conclus**, then we'll reproduce
some of the results in Monti et al. Finally, we'll take a look at a more interesting
example.

#### Plutons ####
The pluton dataset is included in the R package **cluster** and was originally
described in Rousseeuw, Kaufman and Trauwaert, 1996. See the help file for `pluton`
for more information.

```{r pluton, message=FALSE}
library(conclus)
?pluton
dim(pluton)
head(pluton)

gather(pluton, pluton, value) %>%
  ggplot(aes(value)) +
  geom_histogram(fill="blue", bins=15) +
  facet_wrap(~pluton, scales="free")
```
From looking at the histograms, and from reading the help file, it appears the
measurement scales are not really comparable in that, for example, the range of
observable values of Pu238 does not include any of the observable range of values
of Pu239. As such, we might want to scale
each variable before attempting any clustering. Note that this procedure might
or might not aid the analysis.

As noted previously, **conclus** forces the user to think about distance metrics.
(The main reason is that many applications work with mixed continuous and binomial data,
and it's especially important to be clear about how distances between binomial features
are to be defined.) Therefore, we can't just stuff the `pluton` data into `conclus`.
We'll use the `daisy` function from the **cluster** package to create the distance
matrix. First, we'll proceed without rescaling the data.

```{r pluton_daisy}
pd <- daisy(pluton)
ccp <- conclus(pd, K=10)
ggplot(ccp)
```

The `conclus` function defaults to taking 100 subsamples of the data for each choice
of the number of clusters k, (k = 2, ..., 10 in this case). It then computes the
_consensus matrix_ for each value of k: the proportion of times items _i_ and _j_ were
included in the subsample,
and were clustered into the same group. In the figure above, deep blue indicates good
consensus; values near to 1. Peachpuff (that's right, the color is called 'peachpuff')
indicates consensus in the opposite direction; values close to 0 indicating neither
items were in the cluster. Colors between deep blue and peachpuff indicate disagreement
between the different runs of the algorithm.

The reason the panels in the above figure appear approximately block-diagonal is that
the consensus matrices have been clustered themselves, to arrange similar items together.
If perfect consensus were obtained, we'd see a perfectly block-diagonal structure. So
to decide the number of real clusters in the data, we need to choose a value of k for
which we see bunch of blue squares on the diagonal. Maybe 6, 7 or 8 looks ok, but that's
quite a lot of clusters for a fairly small dataset. Let's get some more help

```{r pluton_summary}
sp <- summary(ccp)
sp
```

The output of the summary function shows us the area under the cumulative distribution
function (the AUC of the CDF: compound TLAs!). We are also show summary statistic
Delta, computed as the proportional increase in the AUC resulting from incrementing
k. For example, at k=3, we have (0.642 - 0.473) / 0.475 = 0.358 (adjusting for rounding
error). As we can see,
&Delta;(k) is still quite large, and AUC is rising quite quickly, up till k=5 or 6,
and then they appear to level off.

(Following Monti et al, Section 3.3, &Delta;(k) is actually calculated as the increase
in AUC relative to the maximum AUC for lower values of k.)

The cluster consensus matrix shows the consensus across random subsamples for each
cluster for each value of k. Values of 1 indicate perfect consensus - all subsamples
identified all pairs of observations within that cluster to be in the same cluster.

We can get a better look at the CDF, its AUC and &Delta;(k) by plotting the
summary object.

```{r pluton_summary_plot}
ggplot(sp)
```

Looking at the left-hand panel, the CDFs of the consensus index at each level of k,
we see a big gap between the CDFs for k=2 and k=3, and similarly between k=3 and k=4.
These features are evident in the right-hand panels representing the AUC and &Delta; at
each k. Looking at &Delta; in particular, the number of clusters would look to be maybe 6.

Next, let's redo the whole thing, but scaling the margins of the data first. Note that
you can do that in `daisy` using `stand=TRUE` in the call. However, that uses the mean
to center the data (if you didn't know that, that's the kind of reason **conclus** wants
you to think about distance measures), and I prefer to use the median. So **conclus**
provides function `rscale` as an alternative. Both `daisy` and `rscale` use the median
absolute deviation (mad) as the scale measure. For kicks, we'll also use Manhattan
distance instead of Euclidean distance.

```{r pluton_rscale}
pp <- apply(pluton, 2, rscale)
ccpp <- conclus(daisy(pp, "manhattan"), K=10)
ggplot(ccpp)
```
```{r pluton_rscale_summary}
ggplot(summary(ccpp))
```

From the plots above, there's little difference between those and the versions resulting from
unscaled data and Euclidean distance.

So let's look at what clusters we think we might have discovered, suppose there to be 6.

```{r pluton_pairs, message=FALSE}

pluton$clusters <- membership(ccpp, k=6)
library(GGally) # For ggpairs
pluton$clusters <- factor(pluton$clusters) # to keep ggpairs happy
ggpairs(pluton, columns=1:4, aes(color=clusters), upper="blank")

```
In my humble opinion, that's quite a nice result.

If you've noticed that I've been heavy-handedly stressing the subjectivity involved
in choosing the number of clusters, good! It's deliberate, of course. Having to
apply subjectivity when we don't know what we're looking for is part of the
process.

Finally, notice that earlier in this article, it was stated that there
are lots of clustering algorithms available, but the analysis of the plutons data
didn't even mention which was used! In fact, it used PAM - Partitioning Around Medoids.
That's the default in `conclus` (hint! - see the help file). Other options are
made available, but in the next example, we'll write our own in order to follow
Monti et al closely.

Let's now reproduce some of the results in Monti et al.

#### Uniform1 ####
Monti et al describe various simulated datasets used for evaluating the consensus
clustering approach. The first of these they label _Uniform1_ and is included
in the **conclus** package, having been downloaded from
[the paper's website](http://portals.broadinstitute.org/cgi-bin/cancer/datasets.cgi#87)
(whose URL has been known to change in the past, and don't blame me if it changes in
the future). The data are observations from a 600 dimensional Uniform(0, 1) distribution.

Prior to anything else, we'll need to use the same clustering approach as they used:
hierarchical clustering with average linkage.

If you look at the help file for `conclus`, you'll notice that the default clustering
function is `pamCons`. Others available in the package are `hclustCons` and `fannyCons`.
As the help file states, these are all just simple wrappers around their parent functions,
modified to require precisely 2 arguments, x - a dissimilarity matrix, and k - the number
of clusters to find, and to return nothing other than an integer vector of cluster memberships.

With that in mind, lets write our own function, then pass it into `conclus`, mimicking
the analysis of Monti et al. Also following Monti et al, we'll use 80% subsampling
and 500 samples.

```{r uniform1}
aveHclustCons <- function(x, k){
  cutree(hclust(x, method="average"), k=k)
}

ccu <- conclus(daisy(Uniform1), K=6, cluster=aveHclustCons, subsample=0.8, R=500)
ggplot(ccu, low="white", high="red")
```

Similar to Figure 2 of Monti et al, the consensus matrix for k=3 is largely a  mass of red
mess, with a couple of smaller darker red blocks in the upper right corner (upper left
corner in the original - the matrices are transposed is all).

Let's look at the summary measures and compare them with the top row of Figure 3 in the
original paper.

```{r Uniform1_summary}
ggplot(summary(ccu))
```
The CDFs look similar to in Monti et al, as does the plot of &Delta;(k). We show
the AUC here, which doesn't appear in Monti et al. Let's now create the histogram
of the consensus distribution.

```{r Uniform1_hist}
hist(ccu$M[[2]], col="green") # 2 because k=2 is in the first, so k=3 in the second
```
Again, the plot looks similar to the version in Monti et al.

Let's now move on to the Gaussian3 example.

#### Gaussian3 ####
Following Monti et al, we'll use the same approach with Gaussian3 as we followed
with Uniform1.

```{r Gaussian3}
ccg <- conclus(daisy(Gaussian3), K=6, cluster=aveHclustCons, subsample=0.8, R=500)
ggplot(ccg, low="white", high="red")
```
For k=3, we clearly see the separation of the data into the known 3 groups.

```{r Gaussian3_summary}
ggplot(summary(ccg))
```
Again, the plot of &Delta;(k), and of the CDFs, look like those in Monti et al. Notice that
the CDF for k=3 is a perfect step function. In this case, the algorithm has achieved perfect
consensus with 3 clusters.

```{r Gaussian_summary_text}
summary(ccg)
```
From the textual summary, we see that with k=3, there is 100% consensus (i.e. all
(_i_, _j_) pairs are clustered together in every subsample they both appear in), so
the CDF appears to be correct.

#### St Jude's leukemia data ####
The final example is intended to be a little more interesting and will go beyond
outlining the main features of **conclus**, hopefully finding some insights into
the data.

The St Jude leukemia data consist of 983 measurements taken on each of 246 patients.
There's no particular reason to cluster features - we can just as well cluster
patients (and doing so will be faster).

We are told that there are 6 distinct groups in the data. With such a large number
of dimensions, it will be pretty fruitless to look at simple plots, so we'll
dive in and do the clustering.
We'll also use this example to show off the parallelization
of `conclus` (**note** though that using more than a single core (the default) does not
necessarily speed things up).

```{r stjude_conclus}

sstjude <- apply(stjude, 2, rscale)
dj <- daisy(t(sstjude))

ccj <- conclus(dj, K=10, ncores=7)
s <- summary(ccj)
s
ggplot(s)
```
From the output, 6 (or maybe 5 or 7) clusters looks reasonable. So, success!

It would be nice, though, to see if we can figure out how the features are defined.
That's beyond the scope of this article, though. Good luck! And have fun with **conclus**.

